Explicit B-spline regularization in diffeomorphic image registration

### 2013-10-19 paper.tex
ENH:  adding some mnor edits.

\begin{abstract}
Important methodological developments in the evolution of image registration algorithms include those
in which the correspondence relationship is characterized by diffeomorphisms.
The  popularity of these approaches is due largely to their
topological properties and success in providing biologically plausible
solutions to small and large deformation
estimation problems. Popular variants of diffeomorphic algorithms include those characterized by
time-varying and constant velocity fields, and symmetrical considerations.
Prior information (i.e. regularization) is used to enforce transform plausibility taking the form of physics-based constraints or through some approximation thereof, e.g. Gaussian smoothing of the vector fields (a la Thirion's Demons \citep{thirion1998}).  In the context of the original Demons' framework, the traditional free-form deformation method can be viewed as a variant in which explicit regularization is achieved through the B-spline basis functions.
This characterization can be used to provide alternative B-spline ``flavored'' diffeomorphic image registration solutions with several advantages which we describe in this work.  Implementation is open source and available through the Insight Toolkit and our Advanced Normalization Tools (ANTs) repository.  A thorough comparative evaluation with the well-known SyN algorithm \citep{avants2008}, implemented within the same framework, and its B-spline analog is performed using open data and open source evaluation tools.
\tiny
\keyFont{ \section{Keywords:}
open science, best practices, reproducibility, software }
\end{abstract}

@@

Establishment of anatomical and functional correspondence
is a crucial step towards gaining insight into certain biological
sciences.  Neuroscience research efforts, such as characterizing
brain morphology, require accurate and robust methods for
producing such mappings.  The
extensive literature detailing methodology is evidence of the rich history of
algorithmic development which continues contemporaneously.
We highlight several key contributions which are particularly relevant to the work presented.

@@

Parallel to this branch of algorithmic progress are the informally
denoted ``dense transforms''
perhaps best exemplified by Thirion's seminal contribution \citep{thirion1998}.
Relationships with earlier elastic \citep{bajcsy1989,gee1993} and fluid \citep{christensen1996} registration methods are detailed in
the works of \cite{bro-nielsen1996} and \cite{pennec1999} who observe that
smoothing via Gaussian convolution, a defining characteristic of Demons,
of the update or displacement
field is a greedy approximation for solving the partial differential equations governing
the physics of an elastic or fluid deformation, respectively.  However, the use of such
approximations entails that physical properties, such as topological
regularity, are no longer guaranteed.

It is interesting to note that within this context, traditional FFD algorithms
can be viewed as a type of fluid-like Demons approach
where, rather than projecting the update field to the space
of regularized fields using Gaussian convolution, gradient fields
are projected to a smooth space characterized by the B-spline
basis functions.  This analogy was hinted at in our earlier work \citep{tustison2009} where we showed that fitting the update field to a B-spline object using a fast approximation routine \citep{tustison2006} is equivalent to a preconditioning of the standard gradient used in gradient descent-based FFD optimization.
This preconditioning is used to mitigate the hemstitching effect induced by the ill-conditioned nature of the traditional gradient-based FFD formulation.
Further details can be gleaned from the ITK implementation of this
earlier work%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'')
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

Continuing from the work of \cite{christensen1996} and subsequent exploration into the mathematical formalisms of diffeomorphisms \citep[e.g.][]{dupuis1998},
the well-known Large Deformation Diffeomorphic Metric Mapping (LDDMM) algorithm
was proposed in \cite{beg2005}.  In contrast to the mapping produced
by \cite{christensen1996}, LDDMM yields the geodesic solution in the space of diffeomorphisms between two images. Since its introduction, LDDMM has
inspired much innovation in the image registration literature.  Applying
the log-Euclidean framework of \cite{arsigny2006}, DARTEL (Diffeomorphic Anatomic Registration using Exponential Lie algebra) uses a constant velocity field parameterization to provide a fast, diffeomorphic alternative \citep{ashburner2007}. Additionally, symmetrical considerations in the velocity field parameterization are discussed in \cite{avants2008} in the context of a cross correlation similarity metric which minimizes the bias of the resulting transformation when selecting the ``fixed'' and ``moving'' images.  A greedy version of this algorithm has proven successful in neuroimaging \citep{klein2009} and pulmonary \citep{murphy2011} applications.

@@

Although many extensions of LDDMM rely
on some form of Gaussian convolution for
regularization \citep[e.g.][]{risser2011}, there has been significant interest in constraining
FFD approaches to the space of diffeomorphisms.
An early attempt reported in \cite{rueckert2006} enforced
diffeomorphic transforms
by concatenating multiple FFD transforms, each of which is constrained
to describe a one-to-one mapping.
\cite{modat2011} incorporated the log-Euclidean
framework for enforcing diffeomorphic transformations and ensuring invertibility.  Similarly, the work of \cite{de-craene2011} provided
a full LDDMM-style algorithm based on B-splines called
{\em temporal free-form deformation} in which the
time-varying velocity field
is modeled using a 4-D B-spline object (3-D + time).  Numerical integration of the mapping propagated within the velocity field yields the transform between parameterized time points.

As alluded to earlier, B-spline approximation can also be used
for regularizing time-varying vector fields in an analogous fashion as
Gaussian convolution.  In this vein and similar to
\cite{de-craene2011}, we reported in \cite{tustison2012a,tustison2012}
the use of an $n$-D + time B-spline object to
represent the characteristic velocity fields.  However, we use the
directly manipulated free-form deformation formulation to improve the solution convergence.  This also facilitates modeling
temporal periodicity and the enforcement of stationary boundaries.
Both this work and our earlier work
\citep{tustison2009} demonstrate that the DMFFD framework is potentially
applicable to the
entire gamut of diffeomorphic registration algorithms and provides
unique ``flavors'' of smoothing possibilities.  However, differences
between the two approaches produce characteristically different
solutions.  In addition to
smoothing kernel differences, Gaussian
convolution tends to ``flatten'' the signal in contrast to an
approximation or fitting of the signal provided by the DMFFD
approach.  Also, whereas
convolution operates entirely within the discrete space,
the fitting process constructs a continuous object prior
to any discrete sampling.  
Interestingly, a similar comparison was made with respect to 
Gaussian derivative estimation  \citep{bourma2007}.  Although 
typically estimated using  truncated, discrete Gaussian 
convolution, an alternative based on 
B-spline approximation demonstrated superior performance with
similar computational cost.  
%Further elaboration on these differences
%are discussed in subsequent sections.

Variants of three popular diffeomorphic algorithms and their DMFFD analogs
(LDDMM,
DARTEL,
 and SyN)
were implemented by the authors
as part of the recent refactoring of the open source
Insight Toolkit (ITK) although related work had been
previously implemented within the popular Advanced
Normalization Tools (ANTs).%
\footnote{
http://stnava.github.io/ANTs
}
Given the popularity and excellent performance of the symmetric normalization (SyN) algorithm, our focus in this work is
its B-spline equivalent which we denote as ``B-spline SyN.''
Evaluation and comparisons of the respective algorithmic instantiations
are performed using the \verb#antsRegistration# program found
in the ANTs repository (also originally developed by the authors).  This permits a direct algorithmic comparison
as potential sources for implementation bias have been removed \citep{tustison2013}.  Additionally, in the spirit of open science,
all text, figures, and
scripts to reproduce the results contained in this work are publicly available online.%
\footnote{
https://github.com/ntustison/BSplineMorphisms
}

@@

Given the spatial domain $\Omega$ of $d-$dimensionality
defined over image $I$, a diffeomorphic
mapping, $\phi$, parameterized over $t \in [0,1]$ transforms the image
$I$ to the target image $J$ using $I\circ\phi(\mathbf{x},1)$ where the
geodesic path $\phi(\mathbf{x},t)$ is described by \citep{beg2005}
\begin{align}
  \inf_{\phi} \left( \int_0^1 \|v(t)\|_L^2 dt +
                     \int_{\Omega} | I \circ \phi^{-1}(\mathbf{x},1) - J |^2 d\Omega
              \right).
  \label{eq:lddmm}
\end{align}
$\phi$ is generated as the solution of the ordinary differential equation
\begin{align}
  \frac{d \phi(\mathbf{x},t)}{dt} = v( \phi(\mathbf{x},t), t ),\,\, \phi(\mathbf{x},0) = \mathbf{Id}
\label{eq:ode}
\end{align}
where $v$ is a time-dependent smooth field (as dictated by the functional norm $L$), $v : \Omega \times t
\rightarrow \mathrm{R}^d$ parameterized by $t \in [0,1]$.  Diffeomorphic mappings
between parameterized time points $\{t_a,t_b\} \in [0,1]$
are obtained from  Eq. (\ref{eq:ode}) through integration of the transport
equation, viz.
\begin{align}
  \label{eq:integral}
\phi(\mathbf{x},t_b) &= \phi(\mathbf{x},t_a) + \int_{t_a}^{t_b} v(\phi(\mathbf{x}), t) dt.
\end{align}
%and $L$ is the functional norm which enforces velocity field regularity.

However,
as pointed out in \cite{avants2008}, implementations of this standard
LDDMM formulation are negatively affected by the lack of optimization
symmetry where arbitrary assignment of fixed and moving images could
lead to different solutions despite the fact that the theoretical
geodesic solution describes the same path forwards and backwards.
This observation led to the symmetric
formulation of Eq. (\ref{eq:lddmm}) found in \cite{avants2008}:
\begin{align}
  \inf_{\phi_1} \inf_{\phi_2} \left(
                     \int_0^{0.5} \left( \|v_1(t)\|_L^2 + \|v_2(1-t)\|_L^2 \right) dt +
                     \int_{\Omega} | I \circ \phi_1^{-1}(\mathbf{x},0.5)
                           - J \circ \phi_2^{-1}(\mathbf{x},0.5) |^2 d\Omega
              \right)
\end{align}
where
\begin{align}
  \frac{d \phi_i(\mathbf{x},t)}{dt} = v_i( \phi_i(\mathbf{x},t), t ),\,\, \phi_i(\mathbf{x},0) = \mathbf{Id}, \,\, i \in \{1,2\}
\end{align}
with extension to arbitrary similarity metric choice, i.e. the second
term is replaced with
\begin{align}
\int_{\Omega} \Pi_{\sim}
                          \left( I \circ \phi_1^{-1}(\mathbf{x},0.5),
                           J \circ \phi_2^{-1}(\mathbf{x},0.5) \right) d\Omega
\end{align}
with a popular choice for $\Pi_{\sim}$ being a local neighborhood cross
correlation \citep{avants2008,avants2011}.
Note that
$t$ is parameterized in opposite directions between $\phi_1$ and $\phi_2$.
A diagrammatic
illustration of the explicit symmetry associated with SyN is shown in
Figure \ref{fig:syn}.

@@

Therefore, in addition to the full-scale
symmetric SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in various applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  In this variant, the discrete time-parameterized
velocity field samples are restricted to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  Since the greedy SyN framework is the focus of the evaluation,
the algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

@@

Although several velocity field regularization
operators have been proposed, many algorithmic instantiations
default to Gaussian smoothing due to its simplicity both in implementation
and complexity terms.  A viable and practical alternative is the DMFFD 
approach based on B-splines for explicit regularization of vector fields.
As presented in \cite{tustison2009}, given the similarity metric
$\Pi_{\sim}$, the $d$-dimensional update field (i.e. preconditioned
gradient field), $\delta v_{i_1,\ldots,i_d}$, is calculated from
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d} = \frac{ \sum_{c=1}^{N_{\Omega}} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c \prod_{j=1}^d B_{i_j}(x_j^c)  
  \cdot \frac{\prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} 
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} }
  {\sum_{c=1}^{N_{\Omega}} \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
\end{align}
where the set of $B(\cdot)$ are the univariate B-spline basis functions for
separately modulating regularity in the solution for each parametric dimension,
$N_\Omega$ is the number of voxels in the reference image domain,
$r$ is the spline order in all dimensions,%
\footnote{
In terms of implementation spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.
}
and $\frac{\partial \Pi_\sim}{\partial \mathbf{x}}$ is the spatial 
similarity metric gradient indexed by $c$.

Similarly, in the case of $d$-dimensional time-parameterized 
diffeomorphic image registration, 
the time-dependent velocity field can be represented
as a $(d + 1)$-dimensional B-spline object
\begin{align}
v(\mathbf{x}, t) = \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d}\sum_{i_t=1}^T v_{i_1,\ldots,i_d,i_t} B_{i_t}(t) \prod_{j=1}^d B_{i_j}(x_j)
\end{align}
where $v_{i_1,\ldots,i_d,i_t}$ is a $(d+1)$-dimensional control point lattice
characterizing the velocity field.  The preconditioned gradient analog of
Equation (\ref{eq:dmffd}) for updating the time-varying velocity field control point lattice
is as follows:
$\delta v_{i_1,\ldots,i_d,i_t}$, given the similarity metric, $\Pi_\sim$,
\begin{align}
\label{eq:tvdmffd}
  \delta v_{i_1,\ldots,i_d,i_t} &= \left( \sum_{c=1}^{N_{\Omega} \times N_t} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c B_{i_t}(t^c)\prod_{j=1}^d B_{i_j}(x_j^c)  \right. \nonumber \\
  &\cdot \left. \frac{B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} \sum_{k_t=1}^{r+1} B_{k_t}^2(t^c)
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} \right) \nonumber \\
  &\cdot\left({\sum_{c=1}^{N_{\Omega}\times N_t}B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)} \right) ^{-1}
\end{align}
which takes into account the temporal locations of the
dense gradient field sampled in $t \in [0,1]$. $N_t$ and $N_\Omega$ are the number
of time point samples and the number of voxels in the reference image domain, respectively.
\footnote{
Additionally, in \cite{tustison2006} it was shown that one could associate each
a sample, $\left( \frac{\partial \Pi}{\partial \mathbf{x}}\right)_c$ with a confidence
weighting.  Thus, in order to enforce stationary boundaries for all described
DMFFD-based diffeomorphic algorithms, we
assign image boundary metric gradients a value of zero with a corresponding
large confidence value.
}

@@

The class {\tt itk::BSplineScatteredDataPointSetToImageFilter} underlies all DMFFD regularization which implements the 
methods described in \cite{tustison2009}.  Although applicable to
various scenarios (e.g. curve and surface estimation), it has
been optimized for imaging applications and multi-threaded
for fast processing on multi-core machines.
Additionally, numerical integration for solving Equations (\ref{eq:ode}) and (\ref{eq:integral}) utilizes Runge-Kutta which
provides a more stable alternative than other
methods \citep{press2007}.  Implementation is provided in the class {\tt itk::TimeVaryingVelocityFieldIntegrationImageFilter}.

>>>>

\begin{abstract}
Important methodological developments in the evolution of image registration algorithms include those
in which the correspondence relationship is characterized by diffeomorphisms.
The  popularity of these approaches is due largely to their
topological properties and success in providing biologically plausible
solutions to deformation and morphological
estimation problems. Popular variants of diffeomorphic algorithms include those characterized by
time-varying and constant velocity fields, and symmetrical considerations.
Prior information (i.e. regularization) is used to enforce transform plausibility taking the form of physics-based constraints or through some approximation thereof, e.g. Gaussian smoothing of the vector fields (a la Thirion's Demons \citep{thirion1998}).  In the context of the original Demons' framework, a free-form deformation methodological variant, the so-called {\it directly manipulated free-form deformation} \citep{tustison2009}, can be viewed as an alternative smoothing approach in which explicit regularization is achieved through 
fast B-spline appromixation.
This characterization can be used to provide alternative B-spline ``flavored'' diffeomorphic image registration solutions with several advantages which we describe in this work.  Implementation is open source and available through the Insight Toolkit and our Advanced Normalization Tools (ANTs) repository.  A thorough comparative evaluation with the well-known SyN algorithm \citep{avants2008}, implemented within the same framework, and its B-spline analog is performed using open data and open source evaluation tools.
\tiny
\keyFont{ \section{Keywords:}
open science, best practices, reproducibility, software }
\end{abstract}

@@

Establishment of anatomical and functional correspondence
is a crucial step towards gaining insight into certain biological
sciences.  Neuroscience research efforts, such as characterizing
brain morphology, require accurate and robust methods for
producing such mappings.  The
extensive literature detailing methodology is evidence of the rich history of
algorithmic development which continues contemporaneously.
We highlight several key historical contributions which are particularly relevant to the work presented.

@@

Parallel to this branch of algorithmic progress are the informally
denoted ``dense transforms''
perhaps best exemplified by Thirion's seminal contribution \citep{thirion1998}.
Relationships with earlier elastic \citep{bajcsy1989,gee1993} and fluid \citep{christensen1996} registration methods are detailed in
the works of \cite{bro-nielsen1996} and \cite{pennec1999} who observe that
smoothing via Gaussian convolution, a defining characteristic of Demons,
of the update or total displacement
field is a greedy approximation for solving the partial differential equations governing
the physics of an elastic or fluid deformation, respectively.  However, the use of such
approximations entails that physical properties, such as topological
regularity, are no longer guaranteed.

It is interesting to note that within this context, traditional FFD algorithms
can be viewed as a type of fluid-like Demons approach
where, rather than projecting the update field to the space
of regularized fields using Gaussian convolution, gradient fields
are projected to a smooth space characterized by the B-spline
basis functions.  This analogy was hinted at in our earlier work \citep{tustison2009} where we showed that fitting the update field to a B-spline object using a fast approximation routine \citep{tustison2006} is equivalent to a preconditioning of the standard gradient used in gradient descent-based FFD optimization.
This preconditioning is used to mitigate the hemstitching effect induced by the ill-conditioned nature of the traditional gradient-based FFD formulation.  We denote
this new FFD variant as directly manipulated free-form deformation (DMFFD) and, 
as part of the ITKv4 refactoring efforts, has been implemented for use with the new
registration framework%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'')
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

Continuing from the work of \cite{christensen1996} and subsequent exploration into the mathematical formalisms of diffeomorphisms \citep[e.g.][]{dupuis1998},
the well-known Large Deformation Diffeomorphic Metric Mapping (LDDMM) algorithm
was proposed in \cite{beg2005}.  In contrast to the mapping produced
by \cite{christensen1996}, LDDMM yields the geodesic solution in the space of diffeomorphisms between two images. Since its introduction, LDDMM has
inspired much innovation in the image registration literature.  Applying
the log-Euclidean framework of \cite{arsigny2006}, DARTEL (Diffeomorphic Anatomic Registration using Exponential Lie algebra) uses a constant velocity field parameterization to provide a fast, diffeomorphic alternative \citep{ashburner2007}. Additionally, symmetrical considerations in the velocity field parameterization are discussed in \cite{avants2008} in the context of a cross correlation similarity metric.
By explicitly symmetrizing the LDDMM formulation, this Symmetric Normalization (SyN) minimizes the bias of the resulting transformation when selecting the ``fixed'' and ``moving'' images.  A greedy version of this algorithm has proven successful in neuroimaging \citep{klein2009} and pulmonary \citep{murphy2011} applications.

@@

Although many extensions of LDDMM rely
on some form of Gaussian convolution for
regularization \citep[e.g.][]{risser2011}, there has been significant interest in constraining
FFD approaches to the space of diffeomorphisms.
An early attempt reported in \cite{rueckert2006} enforced
diffeomorphic transforms
by concatenating multiple FFD transforms, each of which is constrained
to describe a one-to-one mapping.
\cite{modat2011} incorporated the log-Euclidean
framework for enforcing diffeomorphic transformations and ensuring invertibility.  Similarly, the work of \cite{de-craene2011} provided
a full LDDMM-style algorithm based on B-splines called
{\it temporal free-form deformation} in which the
time-varying velocity field
is modeled using a 4-D B-spline object (3-D + time).  Numerical Eulerian integration of the mapping propagated within the velocity field yields the transform between parameterized time points.

As alluded to earlier, B-spline approximation can also be used
for regularizing time-varying vector fields in an analogous fashion as
Gaussian convolution.  In this vein, and similar to
\cite{de-craene2011}, we reported in \cite{tustison2012a,tustison2012}
the use of an $n$-D + time B-spline object to
represent the characteristic velocity fields.  However, we use the
directly manipulated free-form deformation formulation to improve the solution convergence.  This also facilitates modeling
temporal periodicity and the enforcement of stationary boundaries.
Both this work and our earlier work
\citep{tustison2009} demonstrate that the DMFFD framework is potentially
applicable to the
entire gamut of diffeomorphic registration algorithms and provides
unique ``flavors'' of smoothing possibilities.  However, 
the two approaches produce characteristically different
solutions.  In addition to
smoothing kernel differences, Gaussian
convolution tends to ``flatten'' the signal in contrast to an
approximation or fitting of the signal provided by the DMFFD
approach.  Also, whereas
Gaussian convolution operates entirely within discretized space,
the B-spline approximation routine constructs a continuous object prior
to any voxelwise reconstruction of the sampled fields.  
Interestingly, a similar comparison was made with respect to 
Gaussian derivative estimation  \citep{bourma2007}.  Although 
typically estimated using  truncated, discrete Gaussian 
convolution, an alternative based on 
B-spline approximation demonstrated superior performance with
similar computational cost.  
%Further elaboration on these differences
%are discussed in subsequent sections.

Variants of three popular diffeomorphic algorithms and their DMFFD analogs
(LDDMM,
DARTEL,
 and SyN)
were implemented by the authors
as part of the recent refactoring of the open source
Insight Toolkit (ITK) although related work had been
previously implemented within the popular Advanced
Normalization Tools (ANTs).%
\footnote{
http://stnava.github.io/ANTs
}
Given the popularity and excellent performance of the greedy variant of SyN algorithm, our evaluation focus in this work is
its B-spline analog which we denote as ``B-spline SyN'' or ``DMFFD SyN.''
Evaluation and comparisons of the respective algorithmic instantiations
are performed using the \verb#antsRegistration# program found
in the ANTs repository (also originally developed by the authors).  This permits a direct algorithmic comparison
as potential sources for implementation bias have been removed \citep{tustison2013}.  Additionally, in the spirit of open science,
all text, figures, and
scripts to reproduce the results contained in this work are publicly available online.%
\footnote{
https://github.com/ntustison/BSplineMorphisms
}

@@

Given the spatial domain $\Omega$ of $d-$dimensionality
defined over image $I$, a diffeomorphic
mapping, $\phi$, parameterized over $t \in [0,1]$ transforms the image
$I$ to the target image $J$ using $I\circ\phi(\mathbf{x},1)$ where the
geodesic path $\phi(\mathbf{x},t)$ is described by \citep{beg2005}
\begin{align}
  \inf_{\phi} \left( \int_0^1 \|v(t)\|_L^2 dt +
                     \int_{\Omega} | I \circ \phi^{-1}(\mathbf{x},1) - J |^2 d\Omega
              \right).
  \label{eq:lddmm}
\end{align}
$\phi$ is generated as the solution of the ordinary differential equation
\begin{align}
  \frac{d \phi(\mathbf{x},t)}{dt} = v( \phi(\mathbf{x},t), t ),\,\, \phi(\mathbf{x},0) = \mathbf{Id}
\label{eq:ode}
\end{align}
where $v$ is a time-dependent smooth field (as dictated by the functional norm $L$), $v : \Omega \times t
\rightarrow \mathrm{R}^d$.  Diffeomorphic mappings
between parameterized time points $\{t_a,t_b\} \in [0,1]$
are obtained from  Eq. (\ref{eq:ode}) through integration of the transport
equation, viz.
\begin{align}
  \label{eq:integral}
\phi(\mathbf{x},t_b) &= \phi(\mathbf{x},t_a) + \int_{t_a}^{t_b} v(\phi(\mathbf{x}), t) dt.
\end{align}
%and $L$ is the functional norm which enforces velocity field regularity.

However,
as pointed out in \cite{avants2008}, implementations of this standard
LDDMM formulation are negatively affected by the lack of optimization
symmetry where arbitrary assignment of fixed and moving images could
lead to different solutions despite the fact that the theoretical
geodesic solution describes the same path forwards and backwards.
This observation led to the symmetric
formulation of Eq. (\ref{eq:lddmm}) found in \cite{avants2008}:
\begin{align}
  \inf_{\phi_1} \inf_{\phi_2} \left(
                     \int_0^{0.5} \left( \|v_1(t)\|_L^2 + \|v_2(t)\|_L^2 \right) dt +
                     \int_{\Omega} | I \circ \phi_1^{-1}(\mathbf{x},0.5)
                           - J \circ \phi_2^{-1}(\mathbf{x},0.5) |^2 d\Omega
              \right)
\end{align}
where
\begin{align}
  \frac{d \phi_i(\mathbf{x},t)}{dt} = v_i( \phi_i(\mathbf{x},t), t ),\,\, \phi_i(\mathbf{x},0) = \mathbf{Id}, \,\, i \in \{1,2\}
\end{align}
with extension to arbitrary similarity metric choice, i.e. the second
term is replaced with
\begin{align}
\int_{\Omega} \Pi_{\sim}
                          \left( I \circ \phi_1^{-1}(\mathbf{x},0.5),
                           J \circ \phi_2^{-1}(\mathbf{x},0.5) \right) d\Omega
\end{align}
with a popular choice for $\Pi_{\sim}$ being a local neighborhood cross
correlation \citep{avants2008,avants2011}.
Note that
$t$ is parameterized in opposite directions between $\phi_1$ and $\phi_2$.
A diagrammatic
illustration of the explicit symmetry associated with SyN is shown in
Figure \ref{fig:syn}.

@@

Therefore, in addition to the full-scale
symmetric SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in various applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  This is possible since the discrete time-parameterized
velocity field samples are restricted to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  Since the greedy SyN framework is the focus of the evaluation,
the algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

@@

Although several velocity field regularization
operators have been proposed, many algorithmic instantiations
default to Gaussian smoothing due to its simplicity both in implementation
and complexity terms.  As mentioned earlier, 
a viable and practical alternative is the DMFFD 
approach based on B-splines for explicit regularization of vector fields.
As presented in \cite{tustison2009}, given the similarity metric
$\Pi_{\sim}$, the $d$-dimensional update field (i.e. preconditioned
gradient field), $\delta v_{i_1,\ldots,i_d}$, is calculated from
\begin{align}
\label{eq:dmffd}
  \delta v_{i_1,\ldots,i_d} = \frac{ \sum_{c=1}^{N_{\Omega}} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c \prod_{j=1}^d B_{i_j}(x_j^c)  
  \cdot \frac{\prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} 
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} }
  {\sum_{c=1}^{N_{\Omega}} \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
\end{align}
where the set of $B(\cdot)$ are the univariate B-spline basis functions for
separately modulating regularity in the solution for each parametric dimension,
$N_\Omega$ is the number of voxels in the reference image domain,
$r$ is the spline order in all dimensions,%
\footnote{
In terms of implementation spline orders can be specified separately for each dimension but, for simplicity,
we only specify a single spline order.
}
and $\frac{\partial \Pi_\sim}{\partial \mathbf{x}}$ is the spatial 
similarity metric gradient indexed by $c$.

Similarly, in the case of $d$-dimensional time-parameterized 
diffeomorphic image registration, 
the time-dependent velocity field can be represented
as a $(d + 1)$-dimensional B-spline object
\begin{align}
v(\mathbf{x}, t) = \sum_{i_1=1}^{X_1}\ldots\sum_{i_d=1}^{X_d}\sum_{i_t=1}^T v_{i_1,\ldots,i_d,i_t} B_{i_t}(t) \prod_{j=1}^d B_{i_j}(x_j)
\end{align}
where $v_{i_1,\ldots,i_d,i_t}$ is a $(d+1)$-dimensional control point lattice
characterizing the velocity field.  The preconditioned gradient analog of
Equation (\ref{eq:dmffd}) for updating the time-varying velocity field control point lattice is 
\begin{align}
\label{eq:tvdmffd}
  \delta v_{i_1,\ldots,i_d,i_t} &= \left( \sum_{c=1}^{N_{\Omega} \times N_t} \left( \frac{\partial \Pi_\sim}{\partial \mathbf{x}} \right)_c B_{i_t}(t^c)\prod_{j=1}^d B_{i_j}(x_j^c)  \right. \nonumber \\
  &\cdot \left. \frac{B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)}
  {\sum_{k_1=1}^{r+1}\ldots\sum_{k_d=1}^{r+1} \sum_{k_t=1}^{r+1} B_{k_t}^2(t^c)
  \prod_{j=1}^d B_{k_j}^2 (x_j^c)} \right) \nonumber \\
  &\cdot\left({\sum_{c=1}^{N_{\Omega}\times N_t}B_{i_t}^2(t^c) \prod_{j=1}^d B_{i_j}^2 (x_j^c)} \right) ^{-1}
\end{align}
which takes into account the temporal locations of the
dense gradient field sampled in $t \in [0,1]$. $N_t$ and $N_\Omega$ are the number
of time point samples and the number of voxels in the reference image domain, respectively.
\footnote{
Additionally, in \cite{tustison2006} it was shown that one could associate each
a sample, $\left( \frac{\partial \Pi}{\partial \mathbf{x}}\right)_c$ with a confidence
weighting.  Thus, in order to enforce stationary boundaries for all described
DMFFD-based diffeomorphic algorithms, we
assign image boundary metric gradients a value of zero with a corresponding
large confidence value.
}

@@

The class {\tt itk::BSplineScatteredDataPointSetToImageFilter} underlies all DMFFD regularization which is an implementation of the 
methods described in \cite{tustison2006}.  Although applicable to
various scenarios (e.g. curve and surface estimation), it has
been optimized for imaging applications and multi-threaded
for fast processing on suitable machines.
Additionally, numerical integration for solving Equations (\ref{eq:ode}) and (\ref{eq:integral}) utilizes Runge-Kutta which
provides a more stable alternative than other
methods \citep{press2007}.  Implementation is provided in the class {\tt itk::TimeVaryingVelocityFieldIntegrationImageFilter}.

A complete packaging of these classes has been made
available as part of our Advanced Normalization Tools (ANTs) toolkit.%
\footnote{
http://stnava.github.io/ANTs/
}
The \verb#antsRegistration# program%
\footnote{
The long and short command line help menus can be invoked by
the commands `{\tt antsRegistration --help}' and `{\tt antsRegistration -h}', respectively.
}
takes advantage of the enhanced ITKv4
registration framework and was developed by the 
authors to provide a robust and versatile solution for a wide variety of
image registration applications. The basic conceptualization for use is that one 
can
set-up any number of registration ``stages'' with each stage being 
characterized by a specified transform.  For example, a representative 
command call is as follows:
\vspace{2mm}
\lstset{frame = htb,
        framerule = 0.25pt,
        float,
        fontadjust,
        backgroundcolor={\color{listlightgray}},
        basicstyle = {\ttfamily\scriptsize},
        keywordstyle = {\ttfamily\color{listkeyword}\textbf},
        identifierstyle = {\ttfamily},
        commentstyle = {\ttfamily\color{listcomment}\textit},
        stringstyle = {\ttfamily},
        showstringspaces = false,
        showtabs = false
        numbers = none,
        numbersep = 6pt,
        numberstyle={\ttfamily\color{listnumbers}},
        tabsize = 2,
        language=bash,
        floatplacement=!h,
        caption={Representative script containing {\tt antsRegistration} and {\tt antsApplyTransforms} command calls used for evaluation},
        captionpos=b,
        label=listing:script
        }
\begin{lstlisting}
# Register the $fixed and $moving images with initial
# alignment of the centers of intensity followed by 
# the following three stages:
#   rigid -> affine -> B-spline SyN
